\documentclass{ximera}
\input{../preamble.tex}

\author{Anna Davis \and Paul Zachlin} 

\title{About this Project} \license{CC-BY 4.0}


\begin{document}
\begin{abstract}
 Project and contact information.
\end{abstract}
\maketitle

%\section{About this Project}
\emph{Differential Equations, An Interactive Introduction} is the second open textbook we have published.  Most of the content in the book originally appeared in Trench, William F., "Elementary Differential Equations" (2013). Faculty Authored and Edited Books \& CDs. 8. (CC-BY-NC-SA)

We have published this book on the Ximera server in order to incorporate modern tools that can only be accessed online which increase the learning experience.

Our first attempt at creating this resource for Differential Equations course was funded by the Ohio Department of Higher Education Innovation grant (2017-2018).  We worked with Justin Greenly from the Franciscan University of Steubenville, and Felipe L. Martins of Cleveland State University.  We are grateful for their contributions.

Funding and support for the current edition was generously provided by the \href{https://github.com/XimeraProject}{Ximera Project} through the US Department of Higher Ed Open Textbook Pilot Program (2024-present).

For questions and comments about the content, please contact Paul Zachlin at pzachlin@lakelandcc.edu

For questions about XIMERA, please write to ximera@math.osu.edu

%\subsection{Acknowledgements}
We are grateful for the opportunity to share this Differential Equations book with you, and we welcome your feedback.  You may also be interested in our first publication, \href{https://sites.google.com/view/lin-alg-interactive-intro/}{Linear Algebra, An Interactive Introduction}.

\end{document}
